Validation of Visual Statistical Inference, Applied to Linear Models

### 2013-04-11 paper_jasa.tex
working on conclusions

This paper has demonstrated that statistical graphics can be used in statistical inference and validates the lineup protocol proposed by \citet{buja:2009} as a formal framework to accomplish that. For this, specific terminologies were defined. The methods of obtaining $p$-value and estimating power of visual tests were introduced. If it is assumed that observers tend to pick the plot with the strongest signal, which was consistent with the experimental data, the power of visual test can be obtained theoretically.  Interestingly, the theoretical power of visual test is better than that of conventional test in the conditions tested in this paper. The simulation study showed that even in the situation where the conventional test has its assumptions met, visual inference is comparable. The results also suggest that when the underlying assumptions of a conventional test do not hold, visual inference will perform well. Even though a single observer can evaluate a lineup, as suggested in the initial lineup protocol, this work recommends having multiple observers to ensure enough power.

When there is no conventional test available, visual testing procedure can be used due to its few assumptions and non-parametric nature.  Lineups are successfully used to evaluate the power of competing graphical designs in \cite{heike:2012}. This method is also applied in \citet{niladri:2012}. It is hoped that this proves to be extremely valuable in data mining applications, and exploratory analyses, where there are no existing gauges of statistical significance. The lineup protocol operates similarly to general tests, such as two-sided, analysis of variance $F$-tests, and Wilks $\Lambda$ for multivariate analysis of variance. If the null hypothesis is rejected, generally we can say that ``there is something there'' but not specifically what it is in the data that triggers the rejection. Follow-up questions on the reasons provide qualitative insight. Multiple comparisons are often done to refine and understand the results of the general conventional testing, and perhaps some similar approaches might be developed for visual inference.

@@

>>>>

This paper has demonstrated that statistical graphics can be used in statistical inference and validates the lineup protocol proposed by \citet{buja:2009}. Specific terminologies were defined, and methods for obtaining the $p$-value and estimating the power of visual tests were introduced. It was assumed that observers select the plot having the strongest signal in the lineup, in order to calculate the theoretical power,  and the experimental data was consistent with this assumption.  The simulation study showed that the results obtained by visual inference are comparable to those obtained by conventional inference, in the setting where conventional testing can be done. Visual inference is intended to provide valid tests where no conventional test exists, and our experiments in a controlled scenario suggest that it will perform as expected in the intended applications. The power of a visual test increases with the number of observers, which interestingly, leads to a result that the theoretical power of visual test can be better than that of conventional test.

The lineup protocol operates analogously to statistical tests that have broad alternative hypotheses. If the null hypothesis is rejected, generally we can say that ``there is something there'' but not specifically what it is in the data that triggers the rejection. Follow-up questions on the reasons provide qualitative insight. Multiple comparisons are often done to refine and understand the results of the general conventional testing, and perhaps some similar approaches might be developed for visual inference.

@@

Lineups have been successfully applied to evaluate the power of competing graphical designs \citep{heike:2012}, and to detecting signal presence in large $p$, small $n$ data \citep{niladri:2012}. It is hoped that this proves to be  valuable in data mining applications, and exploratory analyses, where there are no existing gauges of statistical significance.
